What is an Ancestral Recombination Graph?

### 2022-01-30 paper.tex
Add new introduction

@@

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately, however, there is substantial confusion over
terminology and little agreement on shared standards for data interchange,
significantly hampering progress in this important area.
This confusion and lack of agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a representation of the coalescent with recombination
stochastic process as a graph, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individuals monoploid genomes, an edge
% jk-note: this is clumsy
defines a genetic ancestry relationship between two nodes,
and annotations on those edges define the precise patterns of
ancestry in intervals along the genome.
We use this framework to define some classifications and metrics of ARG nodes,
and discuss the properties of some recently developed inference methods.
\end{abstract}

@@

\section*{Introduction}

% HYW: this is a rough first pass, aiming to be concise. We should flash out the "recombination" aspect of ARGs here a bit, I think

For a long time, the Ancestral Recombination Graph (ARG) has been an object of
fascination to those interested in theoretical evolutionary genetics, as it
captures all knowable genetic history of a collection of individual genomes.
Recently advances in our ability to infer ARGs has led to a resurgence of
interest in this field. However, these have also given rise to certain
terminological confusion, which we aim to clear up in this paper.

One major source of confusion is the use of the term ``ARG" to refer to both a
backwards-in-time generative process (Griffiths) and a structure which can be
created by that process
\citep[e.g.][]{minichiello2006mapping,mathieson2020ancestry}.

@@

In a recombining organism, the genetic ancestry of a longer segment of DNA is
no longer described by a single tree. Instead, ancestral relationships in a
sample can be represented by a sequence of correlated, marginal trees along the
genome. The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a branching-coalescing
stochastic process whose realisations are graphs which into which all of the
marginal ancestral trees are embedded. Coalescences encode common ancestry
as they do in the coalescent, while branching events encode points at which a
marginal tree changes from one position along the genome to the next.
%note-jk: could nitpick this in that diamonds don't generate tree "changes" as such. Maybe tweak this on the next round of revision?

% jk-note: this isn't quite right I think, as it's muddling the distinction between a graph stochastic process
% definition with the graph as the encoding of the marginal trees.
There are several formulations of the ARG, which vary by
the amount of extra information they capture beyond a minimal representation
of all marginal trees. Two prominent examples are the ``big''
ARG~\citep{ethier1990two}, which has a particularly simple specification as a
stochastic process but is computationally infeasible, and the ``little ARG''
traversed by Hudson's algorithm~\citep{hudson1983properties}, which
has a more complex state space and dynamics but can be implemented with
a more manageable runtime and memory footprint. An efficient simulation
implementation has become available only recently \citep{kelleher2016efficient},
prior to which state-of-the-art simulators were based on the so-called sequentially
Markovian coalescent (SMC) \citep{mcvean2005approximating}, which delivers
performance gains by disallowing mergers which would result in trapped
non-ancestral material. An alternative restricted model is
\texttt{ClonalFrame}  \citep{didelot2007inference}, which fixes one distinguished
tree and disallows recombination in any other marginal tree.

@@

In the theoretical literature, ancestral graphs such as the ARG refer to the
ancestral \emph{process} rather than the outcome of a process.

@@

>>>>

\usepackage{subcaption}
\usepackage{tikz}
\usetikzlibrary{calc,positioning}

@@

% JK: this is a rough first pass for a slightly different paper. Needs
% substantial revision.
\begin{abstract}
It has recently become possible to infer genetic ancestry in the presence of
recombination at scale for the first time, enabling many
downstream applications in population and statistical genetics.
Such recombinant genetic ancestry is usually
referred to as an Ancestral Recombination Graph, or ARG.
% Note: slight repetition here with first sentence.
There are now multiple methods that can infer ARGs at a practical scale
and it is therefore vital that these methods can be systematically evaluated
and compared to determine their strengths and weaknesses.
Unfortunately, however, there is substantial confusion over
terminology and little agreement on shared standards for data interchange,
significantly hampering progress in this important area.
This confusion and lack of agreement on basic issues such as what does and
does not constitute an ARG is partly attributable to the historical
development of the term.
Originally rigorously defined as
a representation of the coalescent with recombination
stochastic process as a graph, an ARG is now understood to refer to any
concrete realisation of a recombinant genetic ancestry.
We show that the standard description of an ARG
in terms of common ancestor and recombination events
(inherited from the original stochastic process definition) is
ambiguous, fundamentally limited and an unsuitable basis for data interchange.
We provide a simple alternative definition of an ARG in which a node
corresponds to one of an individuals monoploid genomes, an edge
% jk-note: this is clumsy
defines a genetic ancestry relationship between two nodes,
and annotations on those edges define the precise patterns of
ancestry in intervals along the genome.
We use this framework to define some classifications and metrics of ARG nodes,
and discuss the properties of some recently developed inference methods.
% jk-note: this ends a bit flat - we should say other stuff like we clarify
% how the ARG relates the pedigree, etc, etc. Let's go through again when
% we have a reasonable draft.
\end{abstract}

@@

\section*{Introduction}
Inferring the genetic relationships between sampled genomes in the form of an
evolutionary tree is a necessary prerequisite for many analyses of non (or
rarely) recombining organisms [refs]. The trees capture what is known and
knowable about the genetic ancestry of the sample in an elegant and concise
way, by postulating a set of nodes that are ancestors common to the samples,
and the ancestral relationships between those nodes. A rich literature exists
on analysing the mathematical properties of these trees [cite?], and numerous
methods exist to infer them~\citep{felsenstein2004inferring}. The situation is
quite different in recombining organisms, where the analysis of genetic data
depends on summaries ultimately derived from the trees such as allele
% TODO cite the other papers that preceded ralph2020efficiently here.
frequencies~\citep{ralph2020efficiently}, site frequency spectra, patterns of linkage
disequilibrium~\citep{mcvean2002genealogical} and principal
components~\citep{mcvean2009genealogical}. Genealogical analysis is equally
fundamental to the study of recombining organisms, but profound technical
difficulties have made the routine inference of recombinant genealogies
impossible, and therefore tree-based analysis methods have been largely
unavailable to the fields of population and statistical genetics.

Recent breakthroughs large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2021unified}
and data representation~\citep{kelleher2016efficient}
have raised the genuine prospect of genealogical analysis becoming a standard part
of the population and statistical genetics toolkit.
Applications using these inferred ancestries as input have
begun to appear~\citep{osmond2021estimating} and many more are sure to
follow~\citep{harris2019database}. This vibrant research area, however,
has a significant difficulty, which, if not addressed, will substantially hamper
progress. We currently lack a well-defined data-model and shared terminology
to discuss recombining genealogies,
leading to basic errors in statements about what different inference methods
produce as output (see the XXX section below), as well as presenting serious
difficulties in either comparing the outputs of different methods or
(from a users perspective) using multiple inference methods in an analysis.

In this paper we [turn into proper round up paragraph later]
\begin{enumerate}
\item Discuss the classical Ancestral Recombination Graph formulation,
and show how it derives from (and is limited by) the coalescent with
recombination.
\item Suggest a definition of ARGs that is free of these limitations,
where the formulation is derived from a diploid pedigree rather than
the coalescent approximations and spatial annotations are attached to
edges rather than nodes.
\item Suggest a classification of ARG nodes that helps us to understand
the properties of the structures inferred by different methods, and illustrate
with some examples.
\end{enumerate}

@@

In a recombining organism, the genetic ancestry of a longer segment of DNA is
no longer described by a single tree. Instead, ancestral relationships in a
sample can be represented by a sequence of correlated, marginal trees along the
genome. The Ancestral Recombination Graph (ARG) was introduced by
Griffiths~\citep{griffiths1991two,griffiths1997ancestral} as a branching-coalescing
stochastic process whose realisations are graphs which into which all of the
marginal ancestral trees are embedded. Coalescences encode common ancestry
as they do in the coalescent, while branching events encode points at which a
marginal tree changes from one position along the genome to the next.
% jk-note: could nitpick this in that diamonds don't generate tree
% "changes" as such. Maybe tweak this on the next round of revision?

% jk-note: this isn't quite right I think, as it's muddling the
% distinction between a graph stochastic process
% definition with the graph as the encoding of the marginal trees.
There are several formulations of the ARG, which vary by
the amount of extra information they capture beyond a minimal representation
of all marginal trees. Two prominent examples are the ``big''
ARG~\citep{ethier1990two}, which has a particularly simple specification as a
stochastic process but is computationally infeasible, and the ``little ARG''
traversed by Hudson's algorithm~\citep{hudson1983properties}, which
has a more complex state space and dynamics but can be implemented with
a more manageable runtime and memory footprint. An efficient simulation
implementation has become available only recently \citep{kelleher2016efficient},
prior to which state-of-the-art simulators were based on the so-called sequentially
Markovian coalescent (SMC) \citep{mcvean2005approximating}, which delivers
performance gains by disallowing mergers which would result in trapped
non-ancestral material. An alternative restricted model is
\texttt{ClonalFrame}  \citep{didelot2007inference}, which fixes one distinguished
tree and disallows recombination in any other marginal tree.

@@

\begin{figure}
\centering
\begin{tikzpicture}[xscale=0.8, yscale=0.3, node distance=2mm and 20mm]
\tikzset{greynode/.style={circle,fill,inner sep=1},
nodelabel/.style={font=\footnotesize}}

\node (s0) [greynode] {};
\node (s1) [right=of s0,greynode] {};
\node (s2) [right=of s1,greynode] {};
\node [greynode] (s3) at ($(s1) + (0,1)$) {};
\node [greynode] (s4) at ($0.5*(s1) + 0.5*(s2) + (0,2)$) {};
\node [greynode] (s5) at ($0.5*(s0) + 0.5*(s1) + (0,3)$) {};
\node [greynode] (s6) at ($(s1) + (0,5)$) {};

\node [nodelabel,anchor=north west] at ($(s3) + (0,0)$) {$x = 0.3$};
\foreach \u/\lab in {s0/0, s1/1, s2/2} \node[nodelabel,anchor=north] at (\u) {\lab};
\foreach \u/\lab in {s4/4} \node[nodelabel,anchor=south west] at (\u) {\lab};
\foreach \u/\lab in {s5/5} \node[nodelabel,anchor=south east] at (\u) {\lab};
\foreach \u/\lab in {s3/3, s6/6} \node[nodelabel,anchor=south] at (\u) {\lab};

%% Edges
\draw (s1) -- (s3);
\foreach \child/\parent in {s6/s5, s6/s4, s5/s0, s4/s2}
	\draw (\child) -| (\parent);

\draw (s5) -| ($0.5*(s3) + 0.5*(s3-|s5)$) |- (s3);
\draw (s4) -| ($0.5*(s3) + 0.5*(s3-|s4)$) |- (s3);

\end{tikzpicture}
\caption{\label{fig-arg} \textcolor{red}{A picture of a standard ARG for us
to work with for now. We should replace it with something better at some
point.}
 }
\end{figure}

@@

\section*{Discussion}

Discussion points (may or may not be used, certainly would be heavily revised)

\begin{itemize}
\item Terminology based around an inferred ancestry being a ``true''
ARG is impoverished and misleading. The only reasonable interpretation
of a ARG being true or not is whether it's a sample from the coalescent.
Most inference methods (and all that scale well) do not. In any case,
the coalescent is just one model, and a highly idealised one. It is
exceedingly unlikely that the actual ARG describing the ancestry
of any biological population is ``true'' ARG in this sense. The terminology is
impoverished because it reduces the possibilities available
to something either being an ARG or not, whereas there are many different
interesting properties of these structures that we should be able to
discuss and compare.
\end{itemize}
